---
title: "Analysis on Trees Dataset using Descriptive Statistics"
author:
- "Kashika Tyagi"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    df_print: paged
---

## (a) Print the trees dataset
```{r}
library(ggplot2)
library(moments)
trees #loading the dataset
```
```{r}
head(trees) #returns the first 6 rows from the trees dataset
tail(trees) #returns the last 6 rows from the trees dataset
```

```{r}
str(trees) #display the structure of trees dataset
```

## (b) Print the summary measures for each variable in the dataset
```{r}
summary(trees) #prints the summary measures for each variable in the dataset

```

Alternatively, in order to get the count of variables and observations and their appropriate summaries

```{r}
library(Hmisc)
describe(trees)
```


Alternatively, displaying detail summary of each variable
```{r}
trees_girth <- summary(trees$Girth) #display detail summary of the variable Girth 
trees_girth

sprintf("Variance and standard deviation of trees$Girth is %#.2f and %#.2f", var(trees$Girth), sd(trees$Girth)) #display variance and SD

sprintf("Skewness and Kurtosis of trees$Girth is %#.2f and %#.2f", skewness(trees$Girth), kurtosis(trees$Girth)) #display Skewness and Kurtosis
```

```{r}
trees_height  <- summary(trees$Height) #display detail summary of the variable Height 
trees_height

sprintf("Variance and standard deviation of trees$Height is %#.2f and %#.2f", var(trees$Height), sd(trees$Height)) #display variance and SD

sprintf("Skewness and Kurtosis of trees$Height is %#.2f and %#.2f", skewness(trees$Height), kurtosis(trees$Height)) #display Skewness and Kurtosis
```
```{r}
trees_volume  <- summary(trees$Volume) #display detail summary of the variable Volume
trees_volume

sprintf("Variance and standard deviation of trees$Height is %#.2f and %#.2f", var(trees$Volume), sd(trees$Volume)) #display variance and SD

sprintf("Skewness and Kurtosis of trees$Height is %#.2f and %#.2f", skewness(trees$Volume), kurtosis(trees$Volume)) #display Skewness and Kurtosis
```

## (c) Create Histograms and Density Plots

```{r}

hist(trees$Girth) 

#Alternatively, we can plot the histogram more effectively using ggplot package
library(ggplot2)

ggplot(trees, aes(x=Girth)) +
  geom_histogram(binwidth = 1.5, colour="darkblue", fill="steelblue")

#Histogram with mean line

ggplot(trees, aes(x=Girth)) +
 geom_histogram(binwidth = 1.5, colour="black", fill="steelblue") +
  geom_vline(aes(xintercept=mean(Girth)),
            color="red", linetype="dashed", size=1)

#Histogram with density plot

ggplot(trees, aes(x=Girth)) +
  geom_histogram(aes(y=..density..), binwidth = 1.5, colour="black", fill="steelblue") +
    geom_density(alpha=0.2, fill="blue")

```
```{r}
hist(trees$Height) 

#Alternatively, we can plot the histogram more effectively using ggplot package

library(ggplot2)

ggplot(trees, aes(x=Height)) +
  geom_histogram(binwidth = 1.5, colour="darkblue", fill="steelblue")

#Histogram with mean line

ggplot(trees, aes(x=Height)) +
 geom_histogram(binwidth = 1.5, colour="black", fill="steelblue") +
  geom_vline(aes(xintercept=mean(Height)),
            color="red", linetype="dashed", size=1)

#Histogram with density plot

ggplot(trees, aes(x=Height)) +
  geom_histogram(aes(y=..density..), binwidth = 1.5, colour="black", fill="steelblue") +
    geom_density(alpha=0.2, fill="blue")
```
```{r}
hist(trees$Volume) 

#Alternatively, we can plot the histogram more effectively using ggplot package

library(ggplot2)

ggplot(trees, aes(x=Volume)) +
  geom_histogram(binwidth = 1.5, colour="darkblue", fill="steelblue")

#Histogram with mean line

ggplot(trees, aes(x=Volume)) +
 geom_histogram(binwidth = 1.5, colour="black", fill="steelblue") +
  geom_vline(aes(xintercept=mean(Volume)),
            color="red", linetype="dashed", size=1)

#Histogram with density plot

ggplot(trees, aes(x=Volume)) +
  geom_histogram(aes(y=..density..), binwidth = 1.5, colour="black", fill="steelblue") +
    geom_density(alpha=0.2, fill="blue")
```

### Density Plots 

Density Plots are used to view the distribution of each variable

```{r}
#Density plot of Girth

plot(density(trees$Girth)) #kernel density plot

d <- density(trees$Girth)
plot(d, main="Kernel Density of Girth") #filled density plot
polygon(d, col="blue", border="black")

```

```{r}
#Density plot of Height

plot(density(trees$Height)) #kernel density plot

d <- density(trees$Height)
plot(d, main="Kernel Density of Height") #filled density plot
polygon(d, col="blue", border="black")
```

```{r}
#Density plot of Volume

plot(density(trees$Volume)) #kernel density plot

d <- density(trees$Volume)
plot(d, main="Kernel Density of Volume") #filled density plot
polygon(d, col="blue", border="black")
```


## (d) Create Boxplots

Boxplot is used to show the shape of the distribution, its central value, and its variability

Rug is used to see the data points across the box plot; it adds (1-d plot) of the data to the plot.

```{r}
#Boxplot for the variable Girth
boxplot(trees$Girth, horizontal = TRUE, col = c("yellow"),main = "Boxplot of Girth")
rug(trees$Girth, side = 1) #rug is used to see the data points across the box plot

#Boxplot for the variable Height
boxplot(trees$Height, horizontal = TRUE, col = c("green"), main = "Boxplot of Height")
rug(trees$Height, side = 1)

#Boxplot for the variable Volume
boxplot(trees$Volume, horizontal = TRUE, col = c("red"), main = "Boxplot of Volume")
rug(trees$Volume, side = 1)


```

Combined Boxplot of all the 3 variables

The boxplots of Height shows a normal distribution, Girth is slightly positively-skewed and Volume is positively skewed.
```{r}
#Combined Boxplot of all the 3 variables

boxplot(trees$Girth, trees$Height, trees$Volume, horizontal = TRUE, 
        col = c("yellow","green","red"), names= c("Girth","Height","Volume"), main = "Combined Boxplot of all the 3 variables")
```

# (e) Normal Probability Plots

Normal Probability Plots are used to assess whether a dataset is normally distributed or not specifically, to identify outliers or any unusual values.

```{r}
qqnorm(trees$Girth, xlab = "", ylab = "Girth", main = "Girth of trees"); qqline(trees$Girth, col = 2)

qqnorm(trees$Volume, xlab = "", ylab = "Volume", main = "Volume of trees"); qqline(trees$Volume, col = 2)

```

From the visualizations above, we can see that the normal probability distribution plots of Girth and Volume are skewed right i.e., the plot shows a strongly non-linear pattern. Specifically, it shows a quadratic pattern in which all the points are below a reference line drawn between the first and last points.

```{r}
qqnorm(trees$Height, xlab = "", ylab = "Height", main = "Height of trees");
qqline(trees$Height, col = 2)

```

Normal Probability plot of the Height is short-tailed i.e., it shows a non-linear pattern.

